---
title: "Spatial Data (Science)"
subtitle: DPI 610
format:
  html:
    default: true
  pdf:
    default: true
    geometry:
      - margin=1in
    execute:
      warning: false
      message: false
---

```{r}
#| label: setup
#| echo: false

library(sf)
library(tidyverse)
library(geomander)
library(spdep)
library(janitor)

```

# Introduction

Today we are going to practice working with spatial data, specifically:

- Calculating Moran's I
- Working with Intersections and Unions of Spatial Data

# Pre-Lab Exercise (Complete Before Class)

Before coming to the lab session, complete the following:

## Calculating Moran's I
Use the following code and pick a state you are interested in to explore creating a map exhibiting 2020 Presidential Vote Share by precinct. Then calculate Moran's I on a precinct basis. Here's an example for Utah.

```{r moran-i}
#| warning: false
#| message: false
#| echo: false
#| eval: true
#| fig-width: 6
#| fig-height: 4

# Load voting data for Utah
ut <- get_vest("UT", 2020) %>%
  # Calculate dem vote share
  mutate(dvoteshare = pre_20_dem_bid / (pre_20_dem_bid + pre_20_rep_tru),
         # Calculate centroid of each precinct
         cent = st_centroid(geometry))

# Create a neighborhood object for Moran's I
ut_nb <- poly2nb(ut, queen = TRUE)

# Calculate weights for Moran's I
ut_weights <- nb2listw(ut_nb, style = "W", zero.policy=T)

# Perform Moran's I test
moran <- moran.test(ut %>% pull(dvoteshare), 
                    ut_weights,
                    na.action = na.omit)

# Create a spatial polygon object for each county
ut_counties <- ut %>%
  # Group by county and calculate union of precinct geometries
  group_by(COUNTY) %>%
  summarize(geometry = st_union(geometry)) %>%
  # Calculate centroid of each county
  mutate(coords_x = map_dbl(st_point_on_surface(geometry), 1),
         coords_y = map_dbl(st_point_on_surface(geometry), 2))

# Create a map of Utah vote share by precinct and county labels
ggplot(ut) +
  # Fill polygons by vote share
  geom_sf(aes(fill = dvoteshare), color = NA) +
  # Outline counties in black
  geom_sf(data = ut_counties, fill = NA, color = "black", lwd = 0.3) +
  # Add county labels
  geom_text(data = ut_counties, aes(x = coords_x, y = coords_y, label = COUNTY),
            size = 2, color = "black") +
  # Color scale for vote share
  scale_fill_gradient2(
    low = "red",
    mid = "white",
    high = "blue",
    midpoint = 0.5,
    limits = c(0, 1),
    labels = scales::percent,
    name = "Democratic\nVote Share"
  ) +
  theme_void() +
  labs(title = "2020 Presidential Election Results - Utah",
       subtitle = paste("Moran's I:", round(first(moran$estimate),3)))
```

Now, pick a state you are interested in and reproduce this map with county boundaries and labels. The data from `get_vest()` includes a `COUNTY` column. You can dissolve precincts into counties using `group_by()` and `summarize()`, then use `sf::st_point_on_surface()` to position labels at the center of each county.

## Answer

```{r texas-moran-i}
#| label: answer-pre-lab

# Pick a state — change "UT" to any two-letter state abbreviation
tx <- get_vest("TX", 2020) %>%
  mutate(dvoteshare = pre_20_dem_bid / (pre_20_dem_bid + pre_20_rep_tru),
         cent = st_centroid(geometry))

# Calculate Moran's I
tx_nb <- poly2nb(tx, queen = TRUE)
tx_weights <- nb2listw(tx_nb, style = "W", zero.policy = T)

moran <- moran.test(tx %>% pull(dvoteshare),
                    tx_weights,
                    na.action = na.omit)

# Create county boundaries by dissolving precincts
tx_counties <- tx %>%
  group_by(CNTY) %>%
  summarize(geometry = st_union(geometry)) %>%
  mutate(coords_x = map_dbl(st_point_on_surface(geometry), 1),
         coords_y = map_dbl(st_point_on_surface(geometry), 2))

# Build the map with precincts, county boundaries, and county labels
ggplot(tx) +
  geom_sf(aes(fill = dvoteshare), color = NA) +
  geom_sf(data = tx_counties, fill = NA, color = "black", lwd = 0.3) +
  geom_text(data = tx_counties, aes(x = coords_x, y = coords_y, label = CNTY),
            size = 2, color = "black") +
  scale_fill_gradient2(
    low = "red",
    mid = "white",
    high = "blue",
    midpoint = 0.5,
    limits = c(0, 1),
    labels = scales::percent,
    name = "Democratic\nVote Share"
  ) +
  theme_void() +
  labs(title = "2020 Presidential Election Results",
       subtitle = paste("Moran's I:", round(first(moran$estimate), 3)))

```

# Determining Intersecting District Geographies

:::{.callout-note}
## Pacing Guide
**Core (Questions 1-2):** Everyone should complete these during the lab session.
**Stretch (Question 3):** Attempt if time permits --- this is more challenging and open-ended.
:::

## Question One

In the most recent redistricting cycle, the state of Georgia drew new congressional district maps and congressional elections in 2022 were held using these maps. However,  a judge ruled that these maps constituted violations of Section 2 of the Voting Rights Act and ordered the districts to be redrawn. In a special session, the state legislature redrew new districts that were used in the most recent 2024 congressional elections.

Create maps illustrating the vote shares for each of the two congressional maps of Georgia -- the previously proposed congressional districts and the newly implemented ones (`ush` and `ush_remedial`). Include a note for each plot saying the share of majority Democratic districts and the share of majority-Black districts (based on voting age population `f18_pop` and voting age Black population `f18_ap_blk`). A district is "majority-Black" if `f18_ap_blk / f18_pop >= 0.5`.

## Answer One

```{r}
#| label: answer-one

# Step 1: Load the data
ush <- readRDS("data/ush.rds") # Georgia US House (USH) districts spatial data
ush_remedial <- readRDS("data/ush_remedial.rds") # remedial USH

# Step 2: For ush, compute the share of majority-Dem districts (dvoteshare > 0.5)
#         and the share of majority-Black districts (f18_ap_blk / f18_pop >= 0.5)

# class example
ush %>%
  st_drop_geometry() %>%
  summarize(sum(dvoteshare > 0.5) / n()) %>%
  pull()

# my version
ush <- ush %>% # save these vars. for later
  mutate(maj_dem = if_else(dvoteshare > 0.5, 1, 0),
         maj_blk = if_else(f18_ap_blk / f18_pop >= 0.5, 1, 0)) 

ush %>%
  st_drop_geometry() %>%
  summarize(pct_maj_dem = mean(maj_dem),
            pct_maj_blk = mean(maj_blk))

# Step 3: Build a ggplot map of ush using geom_sf(aes(fill = dvoteshare))
#         with scale_fill_gradient2 (red/white/blue). Add subtitle with stats.
ga_districts <- ush %>%
  group_by(district) %>%
  summarize(geometry = st_union(geometry)) %>%
  mutate(coords_x = map_dbl(st_point_on_surface(geometry), 1),
         coords_y = map_dbl(st_point_on_surface(geometry), 2))

ggplot(ush) +
  geom_sf(aes(fill = dvoteshare), color = NA) +
  geom_text(data = ga_districts, 
            aes(x = coords_x, y = coords_y, label = district),
            size = 2, color = "black") +
  scale_fill_gradient2(
    low = "red",
    mid = "white",
    high = "blue",
    midpoint = 0.5,
    limits = c(0, 1),
    labels = scales::percent,
    name = "Democratic\nVote Share"
  ) +
  theme_void() +
  labs(title = "2020 Georgia US House Districts")


# Step 4: Repeat Steps 2-3 for ush_remedial

# Step 5: Combine the two maps side-by-side using library(patchwork): map1 + map2

```

## Question Two

Did the share of majority-minority districts change from the old to the new maps? If so, by how much? Use `f18_pop` and `nh18_wht` to determine the answer. A district is "majority-minority" if `nh18_wht / f18_pop < 0.5`.

## Answer Two

```{r}
#| label: answer-two

# Compute the share of majority-minority districts for ush
# (where non-Hispanic white voting-age population is less than 50%)

# Repeat for ush_remedial

# Compare the two shares

```

## Question Three

Sometimes we want to know how much of one district overlaps with another district. This is particularly useful when comparing old and new versions of districts. To make these comparisons, we often use the `st_intersection()` function. For example, suppose we wanted to see what districts intersect with the old CD 13.

```{r}
#| echo: false

ush %>%
  filter(cd==13) %>%
  st_intersection(ush_remedial) %>%
  st_drop_geometry() %>%
  count(cd_new)
```
We can develop a more useful function that illustrates the amount of overlapping land area contained in one or more districts based on this command.

```{r}
#area function
area_fn <- function(x,y){
  x <- x %>% mutate(area.x = sf::st_area(.),id_x = 1:n())
  y <- y %>% mutate(area.y = sf::st_area(.),id_y= 1:n())
  
  sf::st_intersection(x, y) %>% 
    dplyr::mutate(
      area       = sf::st_area(.),
      proportion_x = as.numeric(area / area.x),
      proportion_y = as.numeric(area / area.y)
    ) %>%
    tibble::as_tibble() %>%
    dplyr::select(
      id_x,id_y,
      proportion_x,
      proportion_y
    ) 
}
```

Evaluating this function reports back the district(s) of interest along with the intersections, as well as the share of the area contained in each intersection (both in terms of the district of interest and in terms of the district it intersects with).

```{r}
#| echo: false

ush %>%
  filter(cd==13) %>%
  area_fn(ush_remedial)
```

Making use of the results of this function, now create a map that illustrates both the old districts where the composition of the district changed meaningfully as well as the new districts that they intersect with. Focus on at least one of CDs 4, 5, 6, 7 or 13. Find a nice way to present this visually!

## Answer Three (Stretch)

```{r}
#| label: answer-three

# Step 1: Use area_fn() to compute overlaps between ush and ush_remedial
#         (remember to call st_make_valid() on both first)

# Step 2: Pick a district of interest (e.g., cd 13) and filter the overlaps

# Step 3: Use ush and ush_remedial to extract the relevant old/new geometries

# Step 4: Build a ggplot with:
#   - geom_sf() for the new (remedial) districts, filled by district number
#   - geom_sf() for the old district boundaries overlaid in red (fill = NA)
#   - Labels for district numbers

```


